\providecommand{\abntreprintinfo}[1]{%
 \citeonline{#1}}
\setlength{\labelsep}{0pt}\begin{thebibliography}{}
\providecommand{\abntrefinfo}[3]{}
\providecommand{\abntbstabout}[1]{}
\abntbstabout{1.45 }

\bibitem[ABOWD 1999a]{ABOWD1999a}
\abntrefinfo{ABOWD}{ABOWD}{1999a}
{ABOWD, G. Classroom 2000: An experiment with the instrumetation of a living
  educational enviornment.
\emph{IBM Systems Journal}, v.~38, n.~4, p.~508--530, 1999a.}

\bibitem[Abowd e Mynatt 2000]{ABOWD_MYNATT2000}
\abntrefinfo{Abowd e Mynatt}{ABOWD; MYNATT}{2000}
{ABOWD, G.~D.; MYNATT, E.~D. Charting past, present, and future research in
  ubiquitous computing.
\emph{ACM Transactions on Computer-Human Interaction}, v.~7, n.~1, p.~29--58,
  2000.}

\bibitem[ALTHUN e ZIMMERMANN 2003]{ALTHUN2003}
\abntrefinfo{ALTHUN e ZIMMERMANN}{ALTHUN; ZIMMERMANN}{2003}
{ALTHUN, B.; ZIMMERMANN, M. Multimedia streaming services: Specification,
  implementation, and retrieval. In: ACM SIGMM INTERNATIONAL WORKSHOP ON 
  MULTIMEDIA INFORMATION RETRIEVAL, 5., 2003, Berkeley, CA, US. 
  \emph{Proceedings...} New York, NY, US: ACM Press, 2003. p.~247--254.}

\bibitem[ANDLEIGH e THAKRAR 1996]{ANDLEIGH1996}
\abntrefinfo{ANDLEIGH e THAKRAR}{ANDLEIGH; THAKRAR}{1996}
{ANDLEIGH, P.~K.; THAKRAR, K. \emph{Multimedia Systems Design}. Upper Saddle
  River, NJ, US: Pretice Hall PTR, 1996.}

\bibitem[APPLE 2004]{APPLE2004}
\abntrefinfo{APPLE}{APPLE}{2004}
{APPLE. \emph{An Introduction to QuickTime - Overview}. 2004.
Dispon{\'\i}vel em:
  $<$http://developer.apple.com/quicktime/qttutorial/overview.html$>$. Acesso
  em: 10 de nov. de 2004.}

\bibitem[Apple Computer, Inc. 2002]{APPLE2002}
\abntrefinfo{Apple Computer, Inc.}{APPLE COMPUTER, INC.}{2002}
{APPLE COMPUTER, INC. \emph{QuickTime Streaming Server, Darwin Streaming
  Server}: Administrator's guide.
[S.l.], 2002.}

\bibitem[BALDOCHI et al. 2002]{BALDOCHI2002b}
\abntrefinfo{BALDOCHI et al.}{BALDOCHI et al.}{2002}
{BALDOCHI, L. et al. Automatic generation of capture and access applications.
  In: BRAZILIAN SYMPOSIUM ON MULTIMEDIA AND HYPERMEDIA SYSTEMS, 8., 2002, 
  Fortaleza, CE, BR. \emph{Proceedings...} Fortaleza, CE, BR: SBC, 2002. p.~100--115.}

\bibitem[BALDOCHI et al. 2003]{BALDOCHI2003}
\abntrefinfo{BALDOCHI et al.}{BALDOCHI et al.}{2003}
{BALDOCHI, L. et al. Building a middleware infrastructure for capture and
  access applications. In: SEMINÁRIO INTEGRADO DE SOFTWARE E HARDWARE, 30., 2003,
  Campinas, SP, BR. \emph{Anais...} [S.l.: s.n.], 2003. p.~299--313.}

\bibitem[BEIGL 2000]{BEIGL2000}
\abntrefinfo{BEIGL}{BEIGL}{2000}
{BEIGL, M. Memoclip: A location-based remembrance appliance.
\emph{Personal Ubiquitous Computing}, v.~4, n.~4, p.~230--233, 2000.}

\bibitem[BIANCHI 1998]{BIANCHI1998}
\abntrefinfo{BIANCHI}{BIANCHI}{1998}
{BIANCHI, M. Autoauditorium: A fully automatic, multi-camera system to
  televisive auditorium presentations. In: DARPA/NIST SMART SPACES TECHNOLOGY
  WORKSHOP, 1998, Gaithersburg, MD, US. \emph{Proceedings...} [S.l.: s.n.], 1998.}

\bibitem[BROWN, BOVEY e CHEN 1997]{BROWN1997}
\abntrefinfo{BROWN, BOVEY e CHEN}{BROWN; BOVEY; CHEN}{1997}
{BROWN, P.~J.; BOVEY, J.~D.; CHEN, X. Context-aware applications: From the
  laboratory to the marketplace.
\emph{IEEE Personal Communications}, v.~4, n.~5, p.~58--64, 1997.}

\bibitem[Cattelan 2003]{CATTELAN2003}
\abntrefinfo{Cattelan}{CATTELAN}{2003}
{CATTELAN, R.~G. iClass: Um sistema para captura e acesso de sessões em
  ambiente educacional.
\emph{Revista Eletrônica de Iniciação Científica da Sociedade Brasileira de
  Computação (REIC-SBC)}, ~19p., Março 2003.}

\bibitem[Cattelan 2004]{CATTELAN2004}
\abntrefinfo{Cattelan}{CATTELAN}{2004}
{CATTELAN, R.~G.
\emph{Construção de aplicações de captura e acesso baseada em recorrência de
  funcionalidades}.
Dissertação (Mestrado) --- ICMC-USP, São Carlos, SP, BR, Fevereiro 2004.}

\bibitem[CHIU et al. 2001]{CHIU2001}
\abntrefinfo{CHIU et al.}{CHIU et al.}{2001}
{CHIU, P. et al. Liteminutes: An internet-based system for multimedia meeting
  minutes. In: INTERNATIONAL WORLD WIDE WEB CONFERENCE, 10., 2001, Hong-Kong,
  China. \emph{Proceedings...} [S.l.: s.n.], 2001. p.~140--149.}

\bibitem[DEY 2001]{DEY2001}
\abntrefinfo{DEY}{DEY}{2001}
{DEY, A. Undersating and using context.
\emph{Personal and Ubiquitous Computing Journal}, v.~1, n.~5, p.~4--7, 2001.}

\bibitem[DEY e ABOWD 1999]{DEY1999}
\abntrefinfo{DEY e ABOWD}{DEY; ABOWD}{1999}
{DEY, A.~K.; ABOWD, G.~D. \emph{Towards a Better Understanding of Context and
  Context-Awareness}.
Atlanta, GA, US, 1999. (Relatório Técnico).}

\bibitem[FAIRHURST 2001]{FAIRHURST2001}
\abntrefinfo{FAIRHURST}{FAIRHURST}{2001}
{FAIRHURST, G. \emph{MPEG-2 Overview}. 2001.
Dispon{\'\i}vel em:
  $<$http://www.erg.abdn.ac.uk/research/future-net/digital-video/mpeg2.html$>$.
  Acesso em: 10 de nov. de 2004.}

\bibitem[Gibbs e Tsichritzis 1995]{GIBBS1995}
\abntrefinfo{Gibbs e Tsichritzis}{GIBBS; TSICHRITZIS}{1995}
{GIBBS, S.~J.; TSICHRITZIS, D.~C. \emph{Authoring Iteractive Multimedia}:
  Objects, environments and frameworks. New York, NY, US: ACM Press, 1995.}

\bibitem[HANDLEY e JACOBSON 1998]{HANDLEY1998}
\abntrefinfo{HANDLEY e JACOBSON}{HANDLEY; JACOBSON}{1998}
{HANDLEY, M.; JACOBSON, V. \emph{SDP: Session Description Protocol}. 
1998. Dispon{\'\i}vel em: $<$http://rfc.sunsite.dk/rfc/rfc2327.html$>$. Acesso
  em: 18 de mar. de 2005.}

\bibitem[HARRISON et al. 1998]{HARRISON1998}
\abntrefinfo{HARRISON et al.}{HARRISON et al.}{1998}
{HARRISON, B. et al. Squeeze me, hold me, tilt me! An exploration of
  manipulative user interfaces. In: CONFERENCE ON HUMAN FACTORS IN COMPUTING 
  SYSTEMS, 1998, Los Angeles, CA, US. \emph{Proceedings...} 
  New York, NY: ACM Press, 1998. p.~17--241.}

\bibitem[Hauptmann 1995]{HAUPTMANN1995}
\abntrefinfo{Hauptmann}{HAUPTMANN}{1995}
{HAUPTMANN, A. Speech recognition in the informedia digital video library: Uses
  and limitations. In: IEEE INTERNATIONAL CONFERENCE ON TOOLS WITH AI, 6., 1995, 
  [S.l.]. \emph{Proceedings...} [S.l.: s.n.], 1995.}

\bibitem[Ishii e Ullmer 1997]{ISHII_ULLMER1997}
\abntrefinfo{Ishii e Ullmer}{ISHII; ULLMER}{1997}
{ISHII, H.; ULLMER, B. Tangible bits: Towards seamless interfaces between
  people, bits and atoms. In: CONFERENCE ON HUMAN FACTORS IN COMPUTING SYSTEMS, 1997, 
  [S.l]. \emph{Proceedings...} New York, NY, US: ACM Press, 1997. p.~234--241.}

\bibitem[JARMASZ e GEORGANAS 1997]{JARMASZ1997}
\abntrefinfo{JARMASZ e GEORGANAS}{JARMASZ; GEORGANAS}{1997}
{JARMASZ, J.~P.; GEORGANAS, N.~D. Designing a distributed multimedia
  synchronization scheduler. In: IEEE INTERNATIONAL CONFERENCE MULTIMEDIA COMPUTING 
  AND SYSTEMS, 1997. Ottawa, Ontario, Canada. \emph{Proceedings...} [S.l.: s.n.], 
  1997. p.~451--457.}

\bibitem[KIDD et al. 1999]{KIDD1999}
\abntrefinfo{KIDD et al.}{KIDD et al.}{1999}
{KIDD, C.~D. et al. The Aware Home: A living laboratory for ubiquitous
  computing research. In: INTERNATIONAL WORKSHOP ON COOPERATIVE BUILDINGS, 2., 
  1999, Pittsburgh, PA, US. \emph{Proceedings...} [S.l.: s.n.], 1999. p.~191--198.}

\bibitem[KOGA e JINZAKI 2000]{KOGA2000}
\abntrefinfo{KOGA e JINZAKI}{KOGA; JINZAKI}{2000}
{KOGA, H.; JINZAKI, A. DV stream transmission over world wide internet by vídeo
  frame recognition. In: INTERNATIONAL PACKET VIDEO WORKSHOP, 10., 2000, [S.l.].
  \emph{Proceedings...} [S.l.: s.n.], 2000.}

\bibitem[LORENZEN 2001]{LORENZEN2001}
\abntrefinfo{LORENZEN}{LORENZEN}{2001}
{LORENZEN, J. \emph{Darwin Streaming Server}: Open, functional and free. 2001.
Dispon{\'\i}vel em:
  $<$http://www.streamingmedia.com/research/center.asp?id=6$>$. Acesso em:
  10 de dez. de 2004.}

\bibitem[LUM e LAU 2002]{LUM2002}
\abntrefinfo{LUM e LAU}{LUM; LAU}{2002}
{LUM, W.~Y.; LAU, F. C.~M. A context-aware decision engine for context
  adaptation.
\emph{IEEE Pervasive Computing}, v.~1, n.~3, p.~41--49, 2002.}

\bibitem[LUTHER 1991]{LUTHER1991}
\abntrefinfo{LUTHER}{LUTHER}{1991}
{LUTHER, A.~C. \emph{Digital Video in the PC Environment}. New York, NY, US:
  McGraw-Hill Book, 1991.}

\bibitem[LUTHER 1994]{LUTHER1994}
\abntrefinfo{LUTHER}{LUTHER}{1994}
{LUTHER, A.~C. \emph{Authoring Iteractive Multimedia}. Cambridge, MA, UK: ACM
  Press, 1994.}

\bibitem[PASCOE 1998]{PASCOE1998}
\abntrefinfo{PASCOE}{PASCOE}{1998}
{PASCOE, J. Adding generic contextual capabilities to wearable computers. In:
  IEEE INTERNATIONAL SYMPOSIUM ON WEARABLE COMPUTERS, 2., 1998, [S.l.].
  \emph{Proceedings...} Washington, DC, US: IEEE Computer Society 1998. v.~2, p.~92--99.}

\bibitem[RICHTER et al. 2001]{RICHTER2001}
\abntrefinfo{RICHTER et al.}{RICHTER et al.}{2001}
{RICHTER, H. et al. Integrating meeting capture within a collaborative team
  environment. In: INTERNATIONAL CONFERENCE ON UBIQUITOUS COMPUTING, 3., 2001, 
  Atlanta, GA, US. \emph{Proceedings...} London, UK: Springer-Verlag, 2001. p.~123--138.}

\bibitem[Rocha 2004]{ROCHA2004}
\abntrefinfo{Rocha}{ROCHA}{2004}
{ROCHA, C. F.~P.
\emph{Anotações com PDAs}: Extensão da Área de escrita e integração com o
  projeto INCA-SERVE.
Dissertação (Mestrado) --- ICMC-USP, São Carlos, Março 2004.}

\bibitem[ROSEMAN e GREENBERG 1996]{ROSEMAN1996}
\abntrefinfo{ROSEMAN e GREENBERG}{ROSEMAN; GREENBERG}{1996}
{ROSEMAN, M.; GREENBERG, S. Teamrooms: Network places for collaboration. In:
  ACM CONFERENCE ON COMPUTER SUPPORTED COOPERATIVE WORK, 1996, [S.l.].
  \emph{Proceedings...} New York, NY, US: ACM Press, 1996. p.~325--333.}

\bibitem[ROSEN e HAUSMAN 2003]{ROSEN2003}
\abntrefinfo{ROSEN e HAUSMAN}{ROSEN; HAUSMAN}{2003}
{ROSEN, A.~L.; HAUSMAN, M. Digital imaging and video: Principles and
  applications.
\emph{Perspectives on Modern Orthopaedics}, v.~11, n.~6, p.~373--379, 2003.}

\bibitem[RYAN, PASCOE e MORSE 1997]{RYAN1997}
\abntrefinfo{RYAN, PASCOE e MORSE}{RYAN; PASCOE; MORSE}{1997}
{RYAN, N.; PASCOE, J.; MORSE, D.~R. Enhanced reality fieldwork: The
  context-aware archaeologist assistant. In: COMPUTER APPLICATIONS AND 
  QUANTITATIVE METHODS IN ARCHAEOLOGY, 1997, [S.l.]. \emph{Proceedings...}
  [S.l.: s.n.], 1997. p.~269--274.}

\bibitem[SALBER, DEY e ABOWD 1998]{SALBER1998}
\abntrefinfo{SALBER, DEY e ABOWD}{SALBER; DEY; ABOWD}{1998}
{SALBER, D.~A.; DEY, A.~K.; ABOWD, G.~D. \emph{Ubiquitous Computing}: Defining
  and HCI research agenda for an emerging interaction paradigm.
Atlanta, GA, US, 1998. (Relatório Técnico).}

\bibitem[Schilit, Golovchinsky e Price 1998]{SCHILIT1998}
\abntrefinfo{Schilit, Golovchinsky e Price}{SCHILIT; GOLOVCHINSKY; PRICE}{1998}
{SCHILIT, B.; GOLOVCHINSKY, G.; PRICE, M. Beyond paper: Supporting active
  reading with free form digital ink annotations. In: CONFERENCE ON HUMAN FACTORS
  IN COMPUTING SYSTEMS, 1998, Los Angeles, CA, US. \emph{Proceedings...} 
  New York, NY, US: ACM Press, 1998. p.~249--256.}  
 
\bibitem[SCHILIT, ADAMS e WANT 1994]{SCHILIT1994}
\abntrefinfo{SCHILIT, ADAMS e WANT}{SCHILIT; ADAMS; WANT}{1994}
{SCHILIT, B.~N.; ADAMS, N.; WANT, R. Context-aware computing applications. In:
  IEEE WORKSHOP ON MOBILE COMPUTING SYSTEMS AND APPLICATIONS, 1., 1994, 
  Santa Cruz, CA, US. \emph{Proceedings...} New York, NY, US: ACM Press, 1994. p.~85--90.}

\bibitem[SCHULZRINNE et al. 1996]{SCHULZRINNE1996}
\abntrefinfo{SCHULZRINNE et al.}{SCHULZRINNE et al.}{1996}
{SCHULZRINNE, H.et al. \emph{RTP: A Transport Protocol for Real-Time Applications}. 
1996. Dispon{\'\i}vel em: $<$http://rfc.sunsite.dk/rfc/rfc1889.html$>$. Acesso
  em: 18 de mar. de 2005.}

\bibitem[SCHULZRINNE, RAO e LANPHIER 1998]{SCHULZRINNE1998}
\abntrefinfo{SCHULZRINNE, RAO e LANPHIER}{SCHULZRINNE; RAO; LANPHIER}{1998}
{SCHULZRINNE, H.; RAO, A.; LANPHIER, R. \emph{Real Time Streaming Protocol (RTSP)}. 
1998. Dispon{\'\i}vel em: $<$http://rfc.sunsite.dk/rfc/rfc2326.html$>$. Acesso
  em: 18 de mar. de 2005.}

\bibitem[SHI et al. 2003]{SHI2003}
\abntrefinfo{SHI et al.}{SHI et al.}{2003}
{SHI, Y. et al. The Smart Classroom: Merging technologies for seamless
  tele-education.
\emph{IEEE Pervasive Computing}, v.~2, n.~2, p.~47--55, 2003.}

\bibitem[SOMMER e ORZESSEK 1998]{SOMMER1998}
\abntrefinfo{SOMMER e ORZESSEK}{SOMMER; ORZESSEK}{1998}
{SOMMER, P.; ORZESSEK, M. \emph{ATM and MPEG-2}: Integrating digital video into
  broad. [S.l.]: Prentice Hall, 1998.}

\bibitem[Standardisation 1996]{MPEG11996}
\abntrefinfo{Standardisation}{STANDARDISATION}{1996}
{STANDARDISATION, I.~O. for. \emph{Short MPEG-1 Description}. 1996.
Dispon{\'\i}vel em:
  $<$http://www.chiariglione.org/mpeg/standards/mpeg-1/mpeg-1.htm$>$. Acesso
  em: 10 de nov. de 2004.}

\bibitem[Standardisation 2000]{MPEG22000}
\abntrefinfo{Standardisation}{STANDARDISATION}{2000}
{STANDARDISATION, I.~O. for. \emph{Short MPEG-2 Description}. 2000.
Dispon{\'\i}vel em:
  $<$http://www.chiariglione.org/mpeg/standards/mpeg-2/mpeg-2.htm$>$. Acesso
  em: 10 de nov. de 2004.}

\bibitem[Standardisation 2002]{MPEG42002}
\abntrefinfo{Standardisation}{STANDARDISATION}{2002}
{STANDARDISATION, I.~O. for. \emph{MPEG-4 Overview}. 2002.
Dispon{\'\i}vel em:
  $<$http://www.chiariglione.org/mpeg/standards/mpeg-4/mpeg-4.htm$>$. Acesso
  em: 10 de nov. de 2004.}

\bibitem[STEINMETZ e NAHRSTEDT 1995]{STEINMETZ1995}
\abntrefinfo{STEINMETZ e NAHRSTEDT}{STEINMETZ; NAHRSTEDT}{1995}
{STEINMETZ, R.; NAHRSTEDT, K. \emph{Multimedia}: Computing, communications and
  applications. Englewood Cliffs, NJ, US: Prentice Hall, 1995.}

\bibitem[STREITZ et al. 1994]{STREITZ1994}
\abntrefinfo{STREITZ et al.}{STREITZ et al.}{1994}
{STREITZ, N. et al. Dolphin: Integrated meeting support across local and remote
  desktop environments and liveboards. In: ACM CONFERENCE ON COMPUTER SUPPORTED
  COOPERATIVE WORK, 1994, Chapel Hill, North Carolina, US. \emph{Proceedings...} 
  New York, NY, US: ACM Press, 1994. p.~345--358.}

\bibitem[TRUONG, ABOWRD e BROTHERTON 2001]{TRUONG2001}
\abntrefinfo{TRUONG, ABOWRD e BROTHERTON}{TRUONG; ABOWRD; BROTHERTON}{2001}
{TRUONG, K.; ABOWD, G.; BROTHERTON, J. Who, what, when, where, how: Design
  issues of capture and access applications.
\emph{International Conference on Ubiquitous Computing}, p.~209--224, 2001.}

\bibitem[Weiser 1991]{WEISER1991}
\abntrefinfo{Weiser}{WEISER}{1991}
{WEISER, M. The computer of the 21st century.
\emph{Scientific American}, v.~265, n.~3, p.~94--104, 1991.}

\bibitem[WILT 2004]{WILT2004}
\abntrefinfo{WILT}{WILT}{2004}
{WILT, A.~J. \emph{The DV, DVCAM, and DVCPRO Formats - tech details, FAQ, and
  links}. 2004.
Dispon{\'\i}vel em: $<$http://www.adamwilt.com/DV-FAQ-tech.html$>$. Acesso em:
  10 de dez. de 2004.}

\bibitem[WU et al. 2001]{WU2001}
\abntrefinfo{WU et al.}{WU et al.}{2001}
{WU, D. et al. Streaming video over the internet: Approaches and directions.
\emph{IEEE Transactions on Circuits and Systems for Video Technology}, v.~11,
  n.~3, p.~282--300, 2001.}

\end{thebibliography}
